---
title: "Analysis scripts"
date: "Sept-Oct, 2024"
author: "Beryl Zhuang"
output:
  rmdformats::robobook:
    highlight: kate
    number_sections: true

---
<style type="text/css">

.book .book-body .page-inner {
  max-width: 1500px;
}

#sidebar {
  background: mistyrose4;
}

#postamble {
  background:#6959cc;
  border-top:solid 10px grey100;
}

.title {
  text-align: center;
  color: #6959cc;
}

.subtitle {
  color: grey100;
}

h2, legend {
  color: snow;
}

h1, h3, h4, h5, h6 {
  color: #6959cc;
}


#content h2 {
    background-color: #6959cc;
}

</style>

```{r setup, echo=FALSE, cache=FALSE, results='hide'}
library(knitr)
library(rmdformats)

## Global options
opts_chunk$set(echo=T,
               collapse=T,
	             cache=FALSE,
               prompt=FALSE,
               tidy=FALSE,
               message=FALSE,
               warning=FALSE,
               cache.lazy = FALSE)

```


# Introduction  
This documentation contains the core analysis script for the manuscript:   
"Discrepancies in readouts between Infinium MethylationEPIC v2.0 and v1.0 reflected in DNA methylation-based tools: implications and considerations for human population epigenetic studies"

Preprint is available at 
https://www.biorxiv.org/content/10.1101/2024.07.02.600461v2



```{r}
# save(EPICv1_beta_matrix,EPICv2_beta_matrix,EPICv1_clocks, EPICv2_clocks,SampleInfo, sample_correlation, file = "example_data.Rdata")

load( "example_data.Rdata")
head(EPICv1_beta_matrix)
head(EPICv2_beta_matrix) # with replicate probes collapsed, and probe names are ILLUMINA NAME, same as in EPICv1

head(EPICv1_clocks)
head(EPICv2_clocks)
head(SampleInfo)
sample_correlation[1:5, 1:5]
```


```{r, message=FALSE, warning=FALSE, echo=F}
library(dplyr)
```


# array-level Spearman correlation

calculate array level spearman correlation among the common probes between EPICv1 and v2
```{r}
identical(rownames(EPICv1_beta_matrix), rownames(EPICv2_beta_matrix)) # must be T
betas <- cbind(EPICv1_beta_matrix, EPICv2_beta_matrix)
colnames(betas) <- c(paste0(colnames(EPICv1_beta_matrix), "_EPICv1"),paste0(colnames(EPICv2_beta_matrix), "_EPICv2")) 
cor_samples <- cor(betas, use = 'pairwise.complete.obs',method = "spearman" )
```


# probe-level Spearman correlations, RMSE, and pooled SD

```{r}
# modifying the rmse function to remove NAs
rmse2 <- function(x, y, na.rm = TRUE){
    res <- sqrt(mean((x-y)^2, na.rm = na.rm))
    return(res)
}

j=1:nrow(EPICv1_beta_matrix)
# EPICv1 v2 beta matrix must have the same row order of common CpGs
probe_spearman_cor <- sapply(j, function(i) cor(EPICv1_beta_matrix[i,], EPICv2_beta_matrix[i,], method = "spearman"))
probe_rmse <- sapply(j, function(i) rmse2(EPICv1_beta_matrix[i,], EPICv2_beta_matrix[i,]))
probe_pooled_SD <- sapply(j, function(i) effectsize::sd_pooled(EPICv1_beta_matrix[i,], EPICv2_beta_matrix[i,]))
```

# technical replicate ICC calculation (array-level)

I'm calculating ICCs for technical replicates within each EPIC version (EPICv1 tech rep ICC)
technical replicates within EPIC version are the raters (column), and each probe is an item (row)  
1. within EPICv1 tech rep
2. between version tech reps (by mean)


We use ICC2 as the results

```{r}
library(psych)


# use one of the two pairs of technical replicate
tech_rep <- c("Sample_1", "Sample_1_rep")

## within EPICv1 technical replicate ICCs
input_data_EPICv1 <- data.frame(v= EPICv1_beta_matrix[, tech_rep[1]],v_rep= EPICv1_beta_matrix[,tech_rep[2]])
within_EPICv1_tech_rep_ICC <- psych::ICC(input_data_EPICv1) #use ICC2

## within EPICv2 technical replicate ICCs
input_data_EPICv2 <- data.frame(v= EPICv2_beta_matrix[, tech_rep[1]],v_rep= EPICv2_beta_matrix[,tech_rep[2]])
within_EPICv2_tech_rep_ICC <- psych::ICC(input_data_EPICv2) #use ICC2

## between EPICv1 and EPICv2 technical replicate ICCs
input_data <- data.frame(v1= EPICv1_beta_matrix[, tech_rep] %>% rowMeans(),v2= EPICv2_beta_matrix[,tech_rep]%>% rowMeans())
between_version_tech_rep_ICC <- psych::ICC(input_data) #use ICC2
```


# hierachical clustering and plots

```{r}
library(pheatmap)
library(RColorBrewer)

color_pDat <- SampleInfo %>% select(Version, Sex)
rownames(SampleInfo) <- SampleInfo$Sample_Name
rownames(color_pDat) <- SampleInfo$Sample_Name
str(color_pDat)
# Specify colors
ann_colors = list(Sex =c(F="#C0C0C0", M ="#696969"),
    Version = c(EPICv1 = "#56BEAC", EPICv2 = "#5E405E"))


pheatmap(sample_correlation , cluster_rows=T, cluster_cols=T,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean", clustering_method = "complete",
         annotation_col = color_pDat,
         annotation_colors = ann_colors,
         show_rownames = F, show_colnames = F,border_color = NA,
         main = '')
```


# Epigenetic age acceleration (EAA) calcualtion


## 1.	EPIC version separate
we separated epigenetic ages of EPICv1 and EPICv2 samples and then calculated EAA independently on EPICv1 and EPICv2 samples by extracting residuals from the linear regression model: Epigenetic age ~ chronological age.


```{r}
# for each version separate
model_EEAA_EPICv1 = lm(Horvath_pantissue~ Age, data = EPICv1_clocks)
acc_residual_EEAA_EPICv1 <- model_EEAA_EPICv1$residuals


model_EEAA_EPICv2 = lm(Horvath_pantissue~ Age, data = EPICv2_clocks)
acc_residual_EEAA_EPICv2 <- model_EEAA_EPICv2$residuals
```

## 2.	EPIC versions combined
we first combined epigenetic ages of EPICv1 and EPICv2 samples and then calculated EAA by extracting residuals from the linear regression model: Epigenetic age ~ chronological age.


```{r}
model_EEAA_combined = lm(Horvath_pantissue~ Age, data = rbind(EPICv1_clocks, EPICv2_clocks))
acc_residual_EEAA_combined <- model_EEAA_combined$residuals
acc_residual_EEAA_combined
```

## 3.	EPIC versions combined and version adjusted: 
we first combined epigenetic ages of EPICv1 and EPICv2 samples and calculated EAA by extracting residuals from the linear regression model: Epigenetic age ~ chronological age + EPIC version. 

```{r}
model_EEAA_combined_adjusted = lm(Horvath_pantissue~ Age + Version, data = rbind(EPICv1_clocks, EPICv2_clocks))
acc_residual_EEAA_combined_adjusted <- model_EEAA_combined_adjusted$residuals
acc_residual_EEAA_combined_adjusted
```


# EPIC version adjusted DunedinPACE, DNAmTL, epiTOC, and DNA methylation-based inflammation, smoking, and alcohol scores


## 1.	EPIC version separate: 
DunedinPACe, DNAmTL, epiTOC, and predictor estimates were calculated for EPICv1 and EPICv2 samples without EPIC version adjustment. These are the calculation from each clock/predictor. No additional calculation needed.
```{r}
input_estimates <- rbind(EPICv1_clocks, EPICv2_clocks) # combine EPICv1 and v2
non_adj_DunedinPACE <- input_estimates$DunedinPACE

non_adj_DunedinPACE
```

## 2.	EPIC versions combined and version adjusted: 
DunedinPACe, DNAmTL, epiTOC, and predictor estimates were first calculated for EPICv1 and EPICv2 samples, and were subsequently adjusted for EPIC versions by regressing out EPIC versions using linear regression model: Rate estimate ~ EPIC version.

```{r}
# DunedinPACE as example
input_estimates <- rbind(EPICv1_clocks, EPICv2_clocks) # combine EPICv1 and v2

model_predictor = lm(DunedinPACE ~ Version, data = input_estimates) # adjust Version
adj_DunedinPACE <- residuals(model_predictor)+mean(input_estimates$DunedinPACE)
adj_DunedinPACE

EPICv1_clocks$adj_DunedinPACE <- adj_DunedinPACE[1:24]
EPICv2_clocks$adj_DunedinPACE <- adj_DunedinPACE[25:48]

```



# Paired t tests and calculate effect size

Using paired t-tests and applying a Bonferroni multiple test correction, we evaluated differences in epigenetic clock estimations and EAAs,  DunedinPACe, DNAmTL, epiTOC, as well as biomarker predictor scores and cell type proportions described below between matched samples assessed on EPICv1 and EPICv2 in a cohort-specific manner. Effect sizes were measured by Cohen’s d. Technical replicates were removed prior to the paired t tests.

```{r}
# remove technical replicates before t-tests
EPICv1_clocks_no_rep <-EPICv1_clocks %>% filter(!grepl("rep", Sample_Name, ignore.case = T))
EPICv2_clocks_no_rep <-EPICv2_clocks %>% filter(!grepl("rep", Sample_Name, ignore.case = T))


## example 1. DunedinPACE without EPIC version adjustment----
t.test(EPICv1_clocks_no_rep$DunedinPACE, EPICv2_clocks_no_rep$DunedinPACE, 
            paired = T, alternative = "two.sided")
library(effsize)
effsize::cohen.d(EPICv1_clocks_no_rep$DunedinPACE,
                    EPICv2_clocks_no_rep$DunedinPACE, 
                    paired=T) # cohens'D must be paired

## example 2. DunedinPACE without EPIC version adjustment----
t.test(EPICv1_clocks_no_rep$adj_DunedinPACE, EPICv2_clocks_no_rep$adj_DunedinPACE, 
            paired = T, alternative = "two.sided")

effsize::cohen.d(EPICv1_clocks_no_rep$adj_DunedinPACE,
                    EPICv2_clocks_no_rep$adj_DunedinPACE, 
                    paired=T)
```

